---
title: "Real World Example"
author: "Thomas Lloyd"
date: "31/05/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Firstly get each clubs number of goals in the season both home and away.
```{r}
# Install dplyr package
library("dplyr")
data = read.csv("~/Downloads/england-premier-league-matches-2018-to-2019-stats.csv", header =TRUE)
home_goals = data %>% group_by(home_team_name) %>%
                summarize(total_home_goals = sum(home_team_goal_count))
away_goals = data %>% group_by(away_team_name) %>%
                summarize(total_away_goals = sum(away_team_goal_count))
goals = merge(home_goals, away_goals, by.x = "home_team_name", by.y = "away_team_name")
goals = goals %>% summarize(team = home_team_name, home_goals = total_home_goals, away_goals = total_away_goals, total_goals = total_home_goals + total_away_goals, avg_gpg = (total_home_goals + total_away_goals)/38)
goals
home_goals = data %>% group_by(home_team_name) %>% filter(Game.Week <= 9) %>%
                summarize(total_home_goals = sum(home_team_goal_count))
away_goals = data %>% group_by(away_team_name) %>% filter(Game.Week <= 9) %>%
                summarize(total_away_goals = sum(away_team_goal_count))
goals1 = merge(home_goals, away_goals, by.x = "home_team_name", by.y = "away_team_name")
goals1 = goals1 %>% summarize(team = home_team_name, home_goals = total_home_goals, away_goals = total_away_goals, total_goals_first_10 = total_home_goals + total_away_goals, avg_gpg_first_10 = (total_home_goals + total_away_goals)/9)
goal1 = goals1 %>% select(team, total_goals_first_10, avg_gpg_first_10)
goals_data = merge(goals, goals1, by.x = "team", by.y = "team")
goals_data
```

```{r}
# Install dplyr package
library("dplyr")
data = read.csv("results.csv", header =TRUE)
data = data %>% select(Season, DateTime, HomeTeam, AwayTeam, FTHG, FTAG) %>% filter(Season == "2018-19")
home_goals = data %>% group_by(HomeTeam, Season) %>%
                summarise(total_home_goals = sum(FTHG), games_home = length(FTHG)) 
colnames(home_goals)[1] = "Team" 
away_goals = data %>% group_by(AwayTeam, Season) %>%
                summarize(total_away_goals = sum(FTAG), games_away = length(FTAG))
colnames(away_goals)[1] = "Team"
goals = merge(home_goals, away_goals)
goals = goals %>% group_by(Team) %>% select(Team, total_home_goals, total_away_goals, games_home, games_away) %>% summarise(goals = total_home_goals + total_away_goals, games = games_home + games_away)
goals_per_season = goals

#goals in early part of season, 2020/21
data = read.csv("results.csv", header =TRUE)
data[["DateTime"]] = as.Date(data[["DateTime"]])
data = data %>% select(Season, DateTime, HomeTeam, AwayTeam, FTHG, FTAG) %>% filter(Season == "2018-19", DateTime <= "2018-10-22")
home_goals = data %>% group_by(HomeTeam, Season) %>%
                summarize(total_home_goals = sum(FTHG), games_home = length(FTHG))
colnames(home_goals)[1] = "Team" 
away_goals = data %>% group_by(AwayTeam, Season) %>%
                summarize(total_away_goals = sum(FTAG), games_away = length(FTAG))
colnames(away_goals)[1] = "Team"
goals = merge(home_goals, away_goals)
goals = goals %>% group_by(Team) %>% select(Team, total_home_goals, total_away_goals, games_home, games_away) %>% summarise(goals = total_home_goals + total_away_goals, games = games_home + games_away)
goals_first_part= goals
```


Now carry out James-Stein estimation.

Firstly I am assuming that each team's goal scoring rate, $X_i$ is distributed as follows $X_i \sim Poisson(\mu_i)$ where $\mu_i$ is the expected goals per game. Therefore across the first 10 matches $\sum_{i=1}^{10}X_i \sim Poisson(10\mu_i)$, which I can assume to be Normally distributed. As the variance is unknown - it can be approximated by $\frac{1}{n}\sum_{j}(X_{ij}-\bar{X_i})^2$.


We now have estimates for the standard deviation, therefore I can translate to obtain vector of 20 variables distributed with required distribution for Stein's paradox.

```{r}
transformed = 2*sqrt(goals_first_part$goals +3/8)
transformed
```
Carry out original James Stein
```{r}
library(pracma)
JS_estimates <- (1-18/(Norm(transformed)^2))*transformed
JS_estimates <- (JS_estimates/2)^2-3/8
JS_estimates <- JS_estimates/9
```

Now carry out James Stein - adapted method
```{r}
adapted_JS_estimates <- mean(transformed) + (1-17/(Norm(transformed-mean(transformed))^2))*(transformed - mean(transformed))
adapted_JS_estimates <- (adapted_JS_estimates/2)^2-3/8
adapted_JS_estimates <- adapted_JS_estimates/9
```

Now compare with actual value, standard estimator and James Stein.

First calculate the risk from each.

```{r}
Norm(goals_first_part$goals/9 -goals_per_season$goals/38)^2
Norm(JS_estimates - goals_per_season$goals/38)^2
Norm(adapted_JS_estimates - goals_per_season$goals/38)^2
plot(goals_per_season$goals/38, rep(2,20),  ylim = c(0.5,3.5), xlim = c(0,3.4), yaxt = "n", ylab =  '', xlab = "Estimates", col = "blue", pch = 15)
points(adapted_JS_estimates, rep(3,20), col = "blue", pch = 16)
points(goals_first_part$goals/9, rep(1,20), col = "blue", pch = 17)
legend(2.6,3.3, legend = c("A.JS", "True Value", "MLE"), pch = c(15,16,17), col = c("blue", "blue", "blue"))
sum(abs(adapted_JS_estimates - goals_per_season$goals/38 ) < abs(goals_first_part$goals/9 - goals_per_season$goals/38))
for (i in 1:20){
    lines(c((goals_per_season$goals/38)[i], adapted_JS_estimates[i]), c(2,3))
    lines(c((goals_per_season$goals/38)[i], (goals_first_part$goals/9)[i]), c(2,1))
}
plot(goals_per_season$goals/38, rep(2,20),  ylim = c(0.5,3.5), xlim = c(0,3), yaxt = "n", ylab =  '', xlab = "Estimates", col = "blue", pch = 15)
points(JS_estimates, rep(3,20), col = "blue", pch = 16)
points(goals_first_part$goals/9, rep(1,20), col = "blue", pch = 17)
sum(abs(JS_estimates - goals_per_season$goals/38 ) < abs(goals_first_part$goals/9 - goals_per_season$goals/38))
for (i in 1:20){
    lines(c((goals_per_season$goals/38)[i], JS_estimates[i]), c(2,3))
    lines(c((goals_per_season$goals/38)[i], (goals_first_part$goals/9)[i]), c(2,1))
}
```

Compute limited translation estimator- and see if any improvement.
```{r}
transformed = 2*sqrt(goals_first_part$goals + 3/8)

limited_estimators <- mean(transformed) + (1-17/(Norm(transformed-mean(transformed))^2))*(transformed - mean(transformed))
for (i in 1:20){
  limited_estimators[i] = min(max(limited_estimators[i], transformed[i] -1),transformed[i] +1)
}
limited_estimators  <- (limited_estimators/2)^2-3/8
limited_estimators<- limited_estimators/9
Norm(goals_first_part$goals/9 -goals_per_season$goals/38)^2
Norm(limited_estimators - goals_per_season$goals/38)^2
plot(goals_per_season$goals/38, rep(2,20),  ylim = c(0.5,3.5), xlim = c(0,3), yaxt = "n", ylab =  '', xlab = "Estimates", col = "blue", pch = 15)
points(limited_estimators, rep(3,20), col = "blue", pch = 16)
points(goals_first_part$goals/9, rep(1,20), col = "blue", pch = 17)
sum(abs(limited_estimators - goals_per_season$goals/38 ) < abs(goals_first_part$goals/9 - goals_per_season$goals/38))
for (i in 1:20){
    lines(c((goals_per_season$goals/38)[i], limited_estimators[i]), c(2,3))
    lines(c((goals_per_season$goals/38)[i], (goals_first_part$goals/9)[i]), c(2,1))
}
limited_estimators
```


Nice tables for output in latex

```{r}
goals1 = goals_first_part %>% group_by(Team, goals)  %>% summarise(avg = goals/9)
goals2 = goals_per_season %>% group_by(Team, goals)  %>% summarise(avg = goals/38)
colnames(goals2)[2:3] = c("Goals in Season", "Season Average GPG")
colnames(goals1)[2:3] = c("Goals in 9 Games", "9 Game Average GPG")
table <- merge(goals1, goals2, by.x = "Team", by.y = "Team") %>% select(Team, "Season Average GPG", "9 Game Average GPG")
table$"MLE SSE" <- abs(goals_first_part$goals/9 - goals_per_season$goals/38)
table$"JS Estimate GPG" <- JS_estimates
table$"JS SSE" <- abs(JS_estimates - goals_per_season$goals/38)
table$"Improvement JS" <- abs(JS_estimates - goals_per_season$goals/38 ) < abs(goals_first_part$goals/9 - goals_per_season$goals/38)
table$"Adapt. JS Estimate GPG" = adapted_JS_estimates
table$"Adapt. JS Error" <- abs(adapted_JS_estimates - goals_per_season$goals/38) 
table$"Improvement Adapt. JS" <- abs(adapted_JS_estimates - goals_per_season$goals/38 ) < abs(goals_first_part$goals/9 - goals_per_season$goals/38)
table

```
error change vector
```{r}
error = table$`MLE SSE` - table$`Adapt. JS Error`
error
```
Risk Table (for untransformed variables)

```{r}
new_means = table$`Season Average GPG`
new_means = 2*sqrt(9*new_means+(3/8)) - 1/(4*sqrt(9*new_means))
#Expected Risk
20 - (17^2)/Norm(new_means - mean(new_means))^2
risk_table= data.frame(Estimator = c("MLE", "JS","Adapt. JS"),"Expected Risk"  = c(20,20 -(18^2)/Norm(new_means)^2, 20 - (17^2)/Norm(new_means - mean(new_means))^2), 'SSE Observed' =c(Norm(new_means-transformed)^2,Norm(new_means-(1-18/(Norm(transformed)^2))*transformed )^2,Norm(new_means -(mean(transformed) + (1-17/(Norm(transformed-mean(transformed))^2))*(transformed - mean(transformed))))^2))
risk_table
```


Now replicate over multiple seasons and see how many teams are improved each season by the adapted James-Stein estimator.

```{r}
library("dplyr")
library(pracma)
season_calculation <- function(seasons, dates){
 table <- data.frame(Season =c(),'MLE SSE' = c(),'JS SSE' = c(), 'JS Improvement' = c(), 'Adapt. JS SSE' = c(), 'Adapt. JS Improvement' = c(), limited_risk = c(), limited_improvement = c())
for (i in 1:length(seasons)){
data = read.csv("results.csv", header =TRUE)
data = data %>% select(Season, DateTime, HomeTeam, AwayTeam, FTHG, FTAG) %>% filter(Season == seasons[i])
home_goals = data %>% group_by(HomeTeam, Season) %>%
                summarise(total_home_goals = sum(FTHG), games_home = length(FTHG)) 
colnames(home_goals)[1] = "Team" 
away_goals = data %>% group_by(AwayTeam, Season) %>%
                summarize(total_away_goals = sum(FTAG), games_away = length(FTAG))
colnames(away_goals)[1] = "Team"
goals = merge(home_goals, away_goals)
goals = goals %>% group_by(Team) %>% select(Team, total_home_goals, total_away_goals, games_home, games_away) %>% summarise(goals = total_home_goals + total_away_goals, games = games_home + games_away)
print(goals)
goals_per_season = goals

#goals in early part of season, 2020/21
data = read.csv("results.csv", header =TRUE)
data[["DateTime"]] = as.Date(data[["DateTime"]])
data = data %>% select(Season, DateTime, HomeTeam, AwayTeam, FTHG, FTAG) %>% filter(Season == seasons[i], DateTime <=dates[i])
home_goals = data %>% group_by(HomeTeam, Season) %>%
                summarize(total_home_goals = sum(FTHG), games_home = length(FTHG))
colnames(home_goals)[1] = "Team" 
away_goals = data %>% group_by(AwayTeam, Season) %>%
                summarize(total_away_goals = sum(FTAG), games_away = length(FTAG))
colnames(away_goals)[1] = "Team"
goals = merge(home_goals, away_goals)
goals = goals %>% group_by(Team) %>% select(Team, total_home_goals, total_away_goals, games_home, games_away) %>% summarise(goals = total_home_goals + total_away_goals, games = games_home + games_away)
print(goals)
goals_first_part= goals

transformed = 2*sqrt(goals_first_part$goals +3/8)
JS_estimates <- (1-18/(Norm(transformed)^2))*transformed
JS_estimates <- (JS_estimates/2)^2-3/8
JS_estimates <- JS_estimates/9
adapted_JS_estimates <- mean(transformed) + (1-17/(Norm(transformed-mean(transformed))^2))*(transformed - mean(transformed))
adapted_JS_estimates <- (adapted_JS_estimates/2)^2-3/8
adapted_JS_estimates <- adapted_JS_estimates/9
JS_risk <- Norm(JS_estimates-goals_per_season$goals/38)^2
JS_improvement = sum(abs(JS_estimates - goals_per_season$goals/38 ) < abs(goals_first_part$goals/9 - goals_per_season$goals/38))
MLE_SSE <-  Norm(goals_first_part$goals/9-goals_per_season$goals/38)^2
adapted_JS_risk <- Norm(adapted_JS_estimates-goals_per_season$goals/38)^2
adapted_JS_improvement <- sum(abs(adapted_JS_estimates - goals_per_season$goals/38 ) < abs(goals_first_part$goals/9 - goals_per_season$goals/38))
limited_estimators <- mean(transformed) + (1-17/(Norm(transformed-mean(transformed))^2))*(transformed - mean(transformed))
for (i in 1:20){
  limited_estimators[i] = min(max(limited_estimators[i], transformed[i] -1),transformed[i] +1)
}
limited_estimators  <- (limited_estimators/2)^2-3/8
limited_estimators<- limited_estimators/9
limited_estimators_risk <- Norm(limited_estimators - goals_per_season$goals/38)^2
limited_improvement <- sum(abs(limited_estimators- goals_per_season$goals/38 ) < abs(goals_first_part$goals/9 - goals_per_season$goals/38))
season = seasons[i]
new_season <- data.frame( season , MLE_SSE, JS_risk, JS_improvement,  adapted_JS_risk, adapted_JS_improvement, limited_estimators_risk, limited_improvement)
table = rbind(table, new_season)
}
return(table)
}
```

use function to carry out task:
```{r}
seasons = c("2015-16", "2016-17", "2017-18","2018-19", "2019-20")
dates = c("2015-10-19", "2016-10-23","2017-10-22","2018-10-22", "2019-10-21")
latex_2 = season_calculation(seasons, dates)
latex_2$season =  c("2015-16", "2016-17", "2017-18","2018-19", "2019-20")
colnames(latex_2) = c('Season', 'MLE SSE', 'JS SSE','JS Improvement', 'Adapt. JS SSE', 'Adapt. JS Improvement', 'Lim. Trans. SSE', 'Lim. Trans. Improvement' )
latex_2
```

latex tables
```{r}
library(xtable)
#Expected risk table
print(xtable(risk_table, type = "latex"), file = "risk_table.tex")
#2018-19 detailed table
print(xtable(table, type = "latex"), file = "2018-19_club.tex")
#season comparisons
print(xtable(latex_2, type = "latex"), file = "season_comparison.tex")
```


